High-Dimensional Regularized Discriminant Analysis

### 2016-01-02 reports/paper-hdrda.Rnw
Updated data classification summary and results

<<setup>>=
setwd("..")
library(ProjectTemplate)
load.project()

# Contaminated Normal Sims
results_contaminated <- lapply(results_block_contaminated[[1]], function(r) {
  with(r, c(error_rates, p=p, mu=mu, contamination_prob=contamination_prob))
})
results_contaminated <- do.call(rbind.data.frame, results_contaminated)
results_contaminated <- tbl_df(results_contaminated)

results_contaminated <- results_contaminated %>%
  group_by(p, contamination_prob, mu) %>%
  summarise_each(funs(mean)) %>%
  arrange(p, contamination_prob, mu)

@@

# Maximum Standard Error for Contaminated Sims
max_sim_se <- with(results_contaminated, max(sqrt(Error * (1 - Error) / 500)))
max_sim_se <- round(max_sim_se, digits=3)

max_se <- 1234567890
@

@@

In Figure \ref{fig:sim-results}, we observe each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the for the Pang, Tong, and
Witten classifiers. For smaller dimensions, the random-forest and HDRDA
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both HDRDA classifiers outperformed all other classifiers, suggesting that their
dimension reduction inherent better captured the classificatory information in
the small samples even in the presence of outliers.

\begin{figure}
<<sim_results, fig.width=8, fig.height=8, dpi=200>>=
p <- ggplot(results_contaminated, aes(x=contamination_prob, y=Error))
p <- p + geom_line(aes(color=Classifier, linetype=Classifier), size=0.65)
p <- p + facet_wrap(~ p, labeller = "label_both")
p <- p + xlab("Contamination Probability") + ylab("Expected Error Rate")
p + theme_bw()
@
\caption{Average classification error rates as a function of the contamination
  probability $\epsilon$. Approximate standard errors were no greater than
  \Sexpr{max_sim_se}.}
\label{fig:sim-results}
\end{figure}

The Pang, Tong, and Witten methods yielded practically the same and consistently
the worst error rates when outliers were present with $\epsilon > 0$, suggesting
that these classifiers were sensitive to outliers. Notice, for example, that
when $p=400$, the error rates of the Pang, Tong, and Witten classifiers
increased dramatically from approximately 19\% when no outliers were present to
approximately 43\% when $\epsilon=0.05$. The sharp increase in average error
rates for these three classifiers continued as $\epsilon$ increased
further. Guo's method always outperformed Pang, Witten, and Tong, but once
outliers were present, the Guo classifier's average error rate was competitive
with neither the HDRDA classifiers nor the random forest.

\begin{figure}
<<sim_results_by_p, fig.width=8, fig.height=8, dpi=200>>=
p2 <- results_contaminated %>%
  mutate(p=as.numeric(as.character(p)),
         contamination_prob=round(contamination_prob, digits=2)) %>%
    filter(contamination_prob %in% c(0.0, 0.1, 0.2, 0.3, 0.4, 0.5)) %>%
    rename(`Contamination Probability`=contamination_prob) %>%
      ggplot(., aes(x=p, y=Error))
p2 <- p2 + geom_line(aes(color=Classifier, linetype=Classifier), size=0.65)
p2 <- p2 + facet_wrap(~ `Contamination Probability`, labeller="label_both")
p2 <- p2 + xlab("Number of Features") + ylab("Expected Error Rate")
p2 + theme_bw()
@
\caption{Average classification error rates as a function of the number of
  features $p$. Approximate standard errors were no greater than
  \Sexpr{max_sim_se}.}
\label{fig:sim-results-by-p}
\end{figure}

In Figure \ref{fig:sim-results-by-p} we again examine the simulation results as
a function of the number of features $p$ for a subset of the values of
$\epsilon$. This set of plots allows us to investigate the effect of feature
dimensionality on classification performance. When no outliers were present
(i.e., $\epsilon=0$), the random-forest classifier was outperformed by all other
classifiers. Furthermore, the HDRDA classifiers were superior in terms of
average error rate in this setting. As $p$ increased, an elevation in average
error rate was expected for all classifiers, but the increase was not observed
to be substantial.

For $\epsilon > 0$, we observed a different behavior in classification
performance. First, the Pang, Tong, and Witten methods along with the random
forest increased in average error rate as $p$ increased. Contrarily, the
performance of the HDRDA and Guo classifiers was hardly affected by $p$.  Also,
as discussed above, the HDRDA classifiers were superior to all other classifiers
for larger values of $p$ with only the random forest outperforming them for
smaller feature dimensions.

@@

For the microarray data sets, we computed error-rate estimates for the competing
classifiers after reducing the dimension of the data sets with feature selection
to $d = 100, 200, \ldots, 1000$, similar to the values of $d$ considered by
\cite{Xu:2009fl}. To compare misclassification rates between different values of
$d$, we applied each value of $d$ considered to the same random allocation of
training and test data. Hence, for a given random partition, the set of features
selected for a smaller value of $d$ is a proper subset of the features selected
for a larger value of $d$. We report the average of the error-rate estimates for
each value of $d$ in Figure \ref{fig:error-rates}. Approximate standard errors
were no greater than \Sexpr{max_se}.

The \emph{HDRDA} classifier was superior in classification performance for the
majority of the simulations. For the Burczysnki, Shipp, and Singh data sets, our
classifier outperformed its competitors for all values of $d$ examined. For the
Nakayama data set, the \emph{HDRDA} classifier \eqref{eq:hdrda-decomposed} was
slightly outperformed by the Dudoit, Pang, and Tong classifiers for small values
of $d$ but attained classification rates superior to its competitors for larger
values of $d$. Furthermore, Figure \ref{fig:error-rates} emphasizes two
strengths of our classifier: reliability of the classifier for relatively small
$d$ and stability of the classifier for large $d$. In contrast to the other
classifiers, the \emph{HDRDA} classifier exhibited less variability in its
misclassification estimates for small values of $d$ and yielded superior
classification accuracy as $d$ increased and was hardly effected by the
inclusion of more variables in the data sets. We attribute the apparent
stability of the error-rate estimates of our classifier for large values of $d$
to its novel dimension-reduction technique.

The average error-rate estimates for the Dudoit, Pang, Tong, and Witten
classifiers were comparable for all values of $d$ examined. Furthermore, the
error-rate estimates of the Dudoit, Pang, and Tong classifiers were
approximately equal throughout our study, and the Witten classifier yielded
approximately equal error-rate estimates for the Burczysnki, Shipp, and Singh
data sets. Furthermore, the error-rate estimates for the Dudoit and Tong
classifiers were nearly identical, which suggests that the improved mean
estimator employed by \cite{Tong:2012hw} provided little improvement to the
classification performance of diagonal linear discriminant analysis.

The classification performance of the the Dudoit, Pang, Tong, and Witten
classifiers for the Shipp and Singh data sets revealed the inability of these
methods to handle larger values of $d$. After closer examination of the
classification studies performed in the authors' original papers, we noticed
that classification accuracy estimates were reported only for small values of
$d$, thereby ignoring the inclusion of a larger number of variables. Hence, we
advise practioners to consider filtering a large number of features before
employing these methods.

The popular Guo classifier is most competitive over large values of $d$ for the
Shipp and Singh data sets. However, for the Burczynski and Nakayama data sets,
the implementation of the Guo method in the {\tt rda} package often classified
every test observation into a single class, thus yielding larger error rates
than its competitors. Furthermore, the Guo classifier exhibited unreliable
classification performance for smaller values of $d$ and required a larger
subset of the features to perform competitively in our study.
\begin{figure}
<<error_rates, fig.width=8, fig.height=8, out.width='.65\\linewidth'>>=
print(NULL)
@
\caption{Averages of the 1000 error-rate estimates of the competing classifiers
  for each data set as a function of $d$. Approximate standard errors are no
  greater than \Sexpr{max_se}.}
\label{fig:error-rates}
\end{figure}
Next, we examine the optimal estimates of the tuning parameters for the
\emph{HDRDA} classifier. For brevity, we provide graphical summaries of the
optimal estimates of $\lambda$ and $\gamma$ in \ref{suppB}. For each of the four
data sets, the distribution of the optimal values of $\lambda$ selected through
cross-validation often had a notable bathtub shape, resembling a Beta
distribution with both shape parameters less than unity. Specifically, the
optimal values of $\lambda$ were typically near 0 or near 1, corresponding to
quadratic and linear discriminant analysis, respectively. For instance, with
smaller values of $d$ applied to the Shipp data set, the range of optimal values
of $\lambda$ was typically no larger than 0.05, indicating there were a large
number of training data sets for which we can safely assume the covariance
matrices are unequal. As $d$ increased, the majority of the optimal values of
$\lambda$ were selected as unity, thereby an increased number of variables
resulted in the \emph{HDRDA} classifier preferring the more parsimonious
assumption of equal covariance matrices. In this instance we can see the
flexibility of our proposed classifier to employ the more realistic assumption
of unequal class covariance matrices, but if optimal, the \emph{HDRDA}
classifier \eqref{eq:hdrda-decomposed} can utilize the more parsimonious
assumption of equal covariance matrices.

>>>>

<<setup>>=
setwd("..")
library(ProjectTemplate)
load.project()
@

<<munge_contaminated_sims>>=
# Contaminated Normal Sims
results_contaminated <- lapply(results_block_contaminated[[1]], function(r) {
  with(r, c(error_rates, p=p, mu=mu, contamination_prob=contamination_prob))
})
results_contaminated <- do.call(rbind.data.frame, results_contaminated)
results_contaminated <- tbl_df(results_contaminated)

results_contaminated <- results_contaminated %>%
  group_by(p, contamination_prob, mu) %>%
  summarise_each(funs(mean)) %>%
  arrange(p, contamination_prob, mu) %>%
  ungroup

@@

# Maximum Standard Error for Contaminated Sims
max_sim_se <- with(results_contaminated, max(sqrt(Error * (1 - Error) / 500)))
max_sim_se <- round(max_sim_se, digits=3)
@

<<munge_microarray_results>>=
# Microarray Classification Results
results_microarray <- lapply(results[[1]], function(r) {
  with(r, c(error_rates, `data_set`=data_set))
})
results_microarray <- do.call(rbind.data.frame, results_microarray)
results_microarray <- tbl_df(results_microarray)

avg_error_microarray <- results_microarray %>%
    group_by(data_set) %>%
    summarise_each(funs(mean)) %>%
    gather(Classifier, Avg_Error, -data_set)

sd_error_microarray <- results_microarray %>%
  group_by(data_set) %>%
  summarise_each(funs(sd)) %>%
  gather(Classifier, SD_Error, -data_set)

# Microarray classification results are presented in a table
# Rows = classifier, columns = data set
# Cells = Error (SD) rounded to 3 digits
results_microarray <- inner_join(avg_error_microarray, sd_error_microarray,
                                 by=c('data_set', 'Classifier')) %>%
  mutate(Table_Cells=sprintf('%.3f (%.3f)', Avg_Error, SD_Error),
         data_set=stringr::str_to_title(data_set),
         Classifier=gsub("_", " ", Classifier)) %>%
  select(-Avg_Error, -SD_Error) %>%
  spread(data_set, Table_Cells)
@

@@

In Figure \ref{fig:sim-results}, we observe each classifier's average
classification error rates for the values of $\epsilon$ and $p$. Unsurprisingly,
the average error rate increased for each classifier as the contamination
probability $\epsilon$ increased regardless of the value of $p$. Sensitivity to
the presence of outliers was most apparent for the for the Pang, Tong, and
Witten classifiers. For smaller dimensions, the random-forest and \emph{HDRDA}
classifiers tended to outperform the remaining classifiers with the random
forest performing best. As the feature dimension increased with $p \ge 300$,
both \emph{HDRDA} classifiers outperformed all other classifiers, suggesting
that their dimension reduction inherent better captured the classificatory
information in the small samples even in the presence of outliers.

\begin{figure}
<<sim_results, fig.width=8, fig.height=8, dpi=200>>=
p <- ggplot(results_contaminated, aes(x=contamination_prob, y=Error))
p <- p + geom_line(aes(color=Classifier, linetype=Classifier), size=0.65)
p <- p + facet_wrap(~ p, labeller = "label_both")
p <- p + xlab("Contamination Probability") + ylab("Average Error Rate")
p + theme_bw()
@
\caption{Average classification error rates as a function of the contamination
  probability $\epsilon$. Approximate standard errors were no greater than
  \Sexpr{max_sim_se}.}
\label{fig:sim-results}
\end{figure}

The Pang, Tong, and Witten methods yielded practically the same and consistently
the worst error rates when outliers were present with $\epsilon > 0$, suggesting
that these classifiers were sensitive to outliers. Notice, for example, that
when $p=400$, the error rates of the Pang, Tong, and Witten classifiers
increased dramatically from approximately 19\% when no outliers were present to
approximately 43\% when $\epsilon=0.05$. The sharp increase in average error
rates for these three classifiers continued as $\epsilon$ increased
further. Guo's method always outperformed Pang, Witten, and Tong, but once
outliers were present, the Guo classifier's average error rate was competitive
with neither the \emph{HDRDA} classifiers nor the random forest.

\begin{figure}
<<sim_results_by_p, fig.width=8, fig.height=8, dpi=200>>=
p2 <- results_contaminated %>%
  mutate(p=as.numeric(as.character(p)),
         contamination_prob=round(contamination_prob, digits=2)) %>%
    filter(contamination_prob %in% c(0.0, 0.1, 0.2, 0.3, 0.4, 0.5)) %>%
    rename(`Contamination Probability`=contamination_prob) %>%
      ggplot(., aes(x=p, y=Error))
p2 <- p2 + geom_line(aes(color=Classifier, linetype=Classifier), size=0.65)
p2 <- p2 + facet_wrap(~ `Contamination Probability`, labeller="label_both")
p2 <- p2 + xlab("Number of Features") + ylab("Average Error Rate")
p2 + theme_bw()
@
\caption{Average classification error rates as a function of the number of
  features $p$. Approximate standard errors were no greater than
  \Sexpr{max_sim_se}.}
\label{fig:sim-results-by-p}
\end{figure}

In Figure \ref{fig:sim-results-by-p} we again examine the simulation results as
a function of the number of features $p$ for a subset of the values of
$\epsilon$. This set of plots allows us to investigate the effect of feature
dimensionality on classification performance. When no outliers were present
(i.e., $\epsilon=0$), the random-forest classifier was outperformed by all other
classifiers. Furthermore, the \emph{HDRDA} classifiers were superior in terms of
average error rate in this setting. As $p$ increased, an elevation in average
error rate was expected for all classifiers, but the increase was not observed
to be substantial.

For $\epsilon > 0$, we observed a different behavior in classification
performance. First, the Pang, Tong, and Witten methods along with the random
forest increased in average error rate as $p$ increased. Contrarily, the
performance of the \emph{HDRDA} and Guo classifiers was hardly affected by $p$.
Also, as discussed above, the \emph{HDRDA} classifiers were superior to all
other classifiers for larger values of $p$ with only the random forest
outperforming them for smaller feature dimensions.

@@

Similar to \cite{Witten:2011kc}, we report the average test error rates obtained
over 250 random training-test partitions in Table \ref{tab:microarray-results}.
Standard deviations of the test error rates are given in the parentheses. The
\emph{HDRDA} and Guo classifiers were superior in classification performance for
the majority of the simulations with the popular Guo classifier being the best
for the Burczysnki data set. Both \emph{HDRDA} methods yielded comparable
results.

<<classification_results, results='asis'>>=
table_caption <- 'The average of the test error rates obtained on gene-expression data sets over 250 random training-test partitions. Standard deviations of the test error rates are given in the parentheses.'

xtable_out <- xtable(results_microarray,
                     caption=table_caption,
                     label='tab:microarray-results')
print(xtable_out, include.rownames=FALSE)
@

The average error-rate estimates for the Pang, Tong, and Witten classifiers were
comparable across all data sets. Furthermore, the average error rates for the
Pang and Tong classifiers were approximately equal throughout our study,
suggesting that the mean and variance estimators used in lieu of the MLEs
provided little improvement to classification accuracies. The random-forest
classifier was competitive only on the Singh data set. This result challenges
the claim of \cite{FernandezDelgado:2014ul} that the random forest is
typically superior. More studies should be performed to validate this statement
in the small-sample, high-dimensional setting.

Finally, the average test error rates for the Witten classifier were the largest
across all data sets. Given that the standard deviations were relatively large,
we hesitate to generalize claims regarding the Witten classifier being the
uniformly worst performer. However, the variability in its classification
accuracy across multiple random partitions of each data set were large enough to
rethink applying it in practice.
